### 2021-07-11 newvariant.tex
stupid tex fix
edit
i think im mostly done

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).

@@

Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
In doing so, we assess 
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.

@@

We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.

@@

The integral of the kernel is referred to as the instantaneous reproduction number:
\begin{equation}
\RR_x(t) = \int K_x(t, \tau) \dtau
\label{eq:instR}
\end{equation}
which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
Both the reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.

Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged.
While the instantaneous generation-interval distribution can change under speed-like changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6 for discussion), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like interventions, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and return to this idea in Section 6.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression at the logit scale) corresponds to the relative speed of the epidemic.
When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
Since we are interested in the relationship between the true relative speed and strength, rather than the inference of relative speed from proportion data, our results are insensitive to the number of strains circulating in the population.

@@

Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
% As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
In all of these cases, we always assume $\rw < \rv$ to reflect the observed relative speed of $\delta = 0.1\pday$.
We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.

@@

We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---that is, whether the incidence of wild type and the variant are increasing or decreasing. 
For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.

@@

Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}; 
therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in trasmission, as argued by \citep{gostic2020practical}.
The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.

In practice, estimating the instantaneous reproduction number is complicated given that it requires incidence of infection.
Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
Here, we choose to focus here on the underlying dynamical mechanisms that may affect inference and assume that the incidence of infection is exactly known.
Assuming that the generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\label{eq:rt}
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.

@@

\begin{figure}[!pht]
\begin{center}
\includegraphics[width=0.9\textwidth]{Rtbias_smooth.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
(C, F, I) Phase planes (time is implicit) showing true (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
Gray lines represent the one-to-one line.
For all simulations, the assumed mean generation interval is equal to the mean generation interval of the wild type (5 days), and the squared coefficient of variation in generation intervals is equal to $\kappa = 1/5$.
The mean generation interval of the variant is equal to 4 days (top row), 5 days (middle row), and 6 days (bottom row).
}
\end{center}
\label{fig:Rtbias}
\end{figure}

@@

\begin{figure}[!pth]
\begin{center}
\includegraphics[width=0.95\textwidth]{control.pdf}
\caption{
\textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
(A--D) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C) and -speed (B,D) interventions when the variant has equal mean generation interval as the wild type (A--B) and when the variant has longer generation intervals than the wild type (C--D).
Vertical lines represent the mean generation interval.
(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ (colored) conditional on the mean generation interval of the variant.
(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ (colored) conditional on the mean generation interval of the variant.
Epidemic strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
The constant-strength intervention assumes that the ratio $\Rv/\Ry{post} = \theta$ remains constant.
The constant-speed intervention assumes that the difference $\rv - \ry{post} = \phi$ remains constant.
Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.
}
\end{center}
\label{fig:strengthspeed}
\end{figure}

@@

We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data.
In doing so, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time.

>>>>

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).

@@

Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.

@@

We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ and $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.

@@

The integral of the kernel is referred to as the instantaneous reproduction number:
\begin{equation}
\RR_x(t) = \int K_x(t, \tau) \dtau
\label{eq:instR}
\end{equation}
which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
Both the instantaneous reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.

Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).

We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression of $p(t)$ at the logit scale) corresponds to the relative speed of the epidemic.
When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
Since we are interested in the relationship between the true relative speed and strength, rather than the inference of relative speed from proportion data, our results are insensitive to the number of strains circulating in the population.

@@

Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
In all of these cases, we always assume $\rw < \rv$ to reflect the observed relative speed of $\delta = 0.1\pday$.
We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.

@@

We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---that is, whether the incidence of wild type and the variant are increasing or decreasing. 
For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant}) should be used with care.

@@

Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}; 
therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.

In practice, estimating the instantaneous reproduction number is complicated given that it requires incidence of infection.
Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
Here, we choose to focus here on the underlying dynamical mechanisms that may affect inference and assume that the incidence of infection is exactly known.
Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\label{eq:rt}
\end{equation}
Under strength-like intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.

@@

\begin{figure}[!pht]
\begin{center}
\includegraphics[width=0.9\textwidth]{Rtbias_smooth.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
(C, F, I) Phase planes (time is implicit) showing true (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
Gray lines represent the one-to-one line.
For all simulations, the assumed mean generation interval is equal to the mean generation interval of the wild type (5 days), and the squared coefficient of variation in generation intervals is equal to $\kappa = 1/5$.
The mean generation interval of the variant is equal to 4 days (top row), 5 days (middle row), and 6 days (bottom row).
}
\label{fig:Rtbias}
\end{center}
\end{figure}

@@

\begin{figure}[!pth]
\begin{center}
\includegraphics[width=0.95\textwidth]{control.pdf}
\caption{
\textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
(A--D) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C) and -speed (B,D) interventions when the variant has equal mean generation interval as the wild type (A--B) and when the variant has longer generation intervals than the wild type (C--D).
Vertical lines represent the mean generation interval.
(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ (colored) conditional on the mean generation interval of the variant.
(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ (colored) conditional on the mean generation interval of the variant.
Epidemic strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
The constant-strength intervention assumes that the ratio $\Rv/\Ry{post} = \theta$ remains constant.
The constant-speed intervention assumes that the difference $\rv - \ry{post} = \phi$ remains constant.
Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.
}
\label{fig:strengthspeed}
\end{center}
\end{figure}

@@

We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data.
In doing so, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time (but not necessarily the relative speed).
